Closing the gap between statistical and scientific workflows for improved forecasts in ecology

### 2025-04-02 forecastflows.tex
forgot to push last change yesterday

These limitations are central to current debates. In climate modeling, increasing model complexity has not necessarily led to reduced uncertainty. For instance, the uncertainty range on the effect of increasing CO2 concentration on temperature have remained largely unchanged \citep{Zelinka2020}. This has driven calls for more rigorous and transparent calibration processes \citep{balaji2022general}. Similar concerns arise in ecology, where strong discrepancies exist between model projections \citep{Cheaib2012}. Some researchers now advocate for the simplification of models, to avoid over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}. If a model becomes too complex, understanding the sources of uncertainty and how they propagate through the model may become nearly impossible.
Each additional process and parameter can increase overall uncertainty to the point where predictions lose their usefulness for decision makers \citep{Saltelli2020}.

% Forecasting - potential improvments (do I say too much ideas)
Applying the workflow to process-based models is a key for opening the black box. It would serve as a guide through the successive steps of model development. In particular, incorporating data simulation would introduce a crucial step between model building and data fitting, ensuring a clear delineation between the two and exposing strong degeneracies in the model design. This approach would force researchers to begin with a simpler version of the model, providing a clear pathway to support---or reject---the additional complexity and new parameters along the iterative development of the model. 
% => quantifying parameter uncertainty
% incorporate additional observations and their respective uncertainties
Model calibration would no longer be just a hidden aspect of model building but a step as crucial as forecasting to gain new ecological insights. 
% It would help properly take into account for any issues regarding non-identifiability.
The workflow could also refocus attention on the research question, defining a clear and limited context in which the model should apply.
Process-based model would once again be a way to answer a research question---whereas today, model simulations have increasingly become a subject of study on their own. An universal workflow provides an opportunity to merge statistical and process-based approaches, integrating mechanistic knowledge and leveraging robust statistical approaches.

@@

% better training! on BOTH estimation AND prediction
% learn how to simulate data will allow students to understad what is the link between the maths (eg a distribution) and the ecological processes that could be a l'oeuvre to generate the observations, make better links bewteen disciplines
% \item estimation: being aware of what is a parameter, and mention uncertainty propagation
% \item forecasting: should be a natural outcome, not a finality

>>>>

These limitations are central to current debates. In climate modeling, increasing model complexity has not necessarily led to reduced uncertainty. For instance, the uncertainty range on the effect of increasing CO2 concentration on temperature have remained largely unchanged \citep{Zelinka2020}. This has driven calls for more rigorous and transparent calibration processes \citep{balaji2022general}. Similar concerns arise in ecology, where strong discrepancies exist between model projections \citep{Cheaib2012}. Some researchers now advocate for the simplification of models, to avoid over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}. If a model becomes too complex, understanding the sources of uncertainty and how they propagate through the model may become nearly impossible.
Each additional process and parameter can increase overall uncertainty to the point where model projections lose their usefulness for decision makers \citep{Saltelli2020}.

% Forecasting - potential improvments (do I say too much ideas)
Applying the workflow to process-based models is a key for opening the black box. It would serve as a guide through the successive steps of model development. In particular, incorporating data simulation would introduce a crucial step between model building and data fitting, ensuring a clear delineation between the two and exposing strong degeneracies in the model design. This approach would force researchers to begin with a simpler version of the model, providing a clear pathway to support---or reject---the additional complexity and new parameters along the iterative development of the model. 
% => quantifying parameter uncertainty
% incorporate additional observations and their respective uncertainties
Model calibration would no longer be just a hidden aspect of model building but a step as crucial as forecasting to gain new ecological insights. 
% It would help properly take into account for any issues regarding non-identifiability, and the model would be tractable! I think it's a word
The workflow could also refocus attention on the research question, defining a clear and limited context in which the model should apply.
Process-based model would once again be a way to answer a research question---whereas today, model simulations have increasingly become a subject of study on their own. An universal workflow provides an opportunity to merge statistical and process-based approaches, integrating mechanistic knowledge and leveraging robust statistical approaches.

@@

% better training! on BOTH estimation AND prediction
% learn how to simulate data will allow students to understand what is the link between the maths (e.g. a distribution) and the ecological processes generating the observations 
% mobilizing their ecological knowledge while thinking of a model
% => foster stronger links between disciplines
% \item estimation: being aware of what is a parameter, and mention uncertainty propagation
% \item forecasting: should be a natural outcome, not a finality
